When to trust phylogeographic inferences?

### 2024-04-05 manuscript/phylogeo.tex
add to ms

As soon as the interaction radius $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}.
Diffusion constants estimated from the population remain accurate with moderate overestimation of small diffusion constants.
But with decreasing diffusion, the population fragments into smaller than smaller subpopulations that don't interact or coalesce.
This manifests in an increasing time to the MRCA, see Fig.~\ref{fig:density_reg}.
Spatial location and the coalescence process are thus strongly coupled, though this coupling manifests itself in population structure rather than estimates of the diffusion coefficients.

@@

The simplest model of this process are FKPP waves that model the population density $\phi(x,t)$ as \citep{fisher_wave_1937,KPP1937,hallatschek_life_2010}

This equation admits solutions with a traveling front, that the population ``invades'' the empty territory with velocity $v = 2\sqrt{D \alpha}$.
Note that the model itself does not have an intrinsic velocity.
The speed at which the population invades space is given as compound of diffusion and accelerated growth in regions of low density and only grows like the square root of diffusion constant.
It is thus not clear a priori how estimates of $D$ will be affected by density regulation, even in a constant habitat.

>>>>

As soon as the interaction radius $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}.
Diffusion constants estimated from the population remain accurate with moderate overestimation of small diffusion constants.
But with decreasing diffusion, the population fragments into smaller than smaller subpopulations that don't interact or coalesce.
This manifests in an increasing time to the MRCA, see Fig.~\ref{fig:density_reg}.
Spatial location and the coalescence process are thus strongly coupled, though this coupling manifests itself in population structure rather than estimates of the diffusion coefficients.
In such a spatially structured population, estimates of historical populations sizes (e.g.~skyline estimates) will be strongly affected.

@@

As discussed above, a static carrying capacity in an equilibrium population has limited effects on estimates of ancestral locations and diffusion coefficients. 
In reality, carrying capacities vary through time. 
Such variation happens on all time scales, ranging from environmental changes over millennia to seasonal fluctuations. 
The circulation of seasonal influenza viruses, for example, varies by many orders of magnitude between summer and winter in temperate climates.
Fluctuations in carrying capacity mean that populations not just spread because individuals move, but also because populations grow in regions where the population density is below the carrying capacity.
In such situations, the accuracy and interpretation of phylogeographic inferences are unclear.

A simple model of the interplay between growth and spatial spread is again given by the Fisher-KPP equation (Eq.~\ref{eq:FKPP}) above. 
This equation admits solutions with a traveling front, that the population ``invades'' the empty territory with velocity $v_{FKPP} = 2\sqrt{D \alpha}$ \citep{fisher_wave_1937,KPP1937,hallatschek_life_2010}.
Note that the model itself does not have an explicit parameter that has dimensions of a velocity.
The speed at which the population invades space is given as compound of diffusion and accelerated growth in regions of low density and only grows like the square root of diffusion constant.

To explore such the result of such changing habitats on phylogeographic inferences, we simulated a habitat with a Gaussian carrying capacity that moves at a constant velocity $v$ along the $x$-axis (Fig.~\ref{fig:traveling_wave}). 
When varying the microscopic dispersal parameter $D$ and the velocity of the carrying capacity $v$, I observe diverse behavior with often dramatically overestimated $\hat{D}$. These behaviors are explained comparing the speed $v_{FKPP}$ at which the population can invade new territory with the external speed $v$, the ratio of which is used to plot the simulation data in Fig.~\ref{fig:traveling_wave}. 
For $v_{FKPP} \ll v$, the population cannot keep up with the moving carrying capacity and is saved from extinction only because selection is soft.
In this regime, the estimated velocity of lineages is below $v$ and both the estimated $v$ and $D$ increase with $v_{FKPP}$. 
Once $v\sim v_{FKPP}$, the estimated lineage velocity coincides with $v_{FKPP}$, while the estimated $D$ peaks at values orders of magnitude above the true value. 
For larger $v_{FKPP}$, the estimated velocity is stable for some time, before increasing again with $v_{FKKP}$ is a sampling dependent manner. 
At larger $v_{FKPP}$, $D$ is eventually underestimates.

In this extreme example of a population dragged along by a moving carrying capacity, estimates of dispersal rates are meaningless, and estimates of velocities are only compatible with the speed of the shifting habitat in a narrow range when the habitat speed matches the FKPP velocity.
Similarly, estimates of ancestral locations are biased to the current position of the habitat.

Less extreme spatio-temporal dynamics of the habitat results in less extreme biases, but still illustrate that caution is required when interpreting phylogeographic inferences in environment that might change between the present and the $T_{MRCA}$ of the population.  
I simulated two additional scenarios: one habitat that expands and contracts to a core region periodically, and one that oscillates back and forth between two regions with a homogeneous intermediate state.

\begin{figure}
    \includegraphics[width=0.48\textwidth]{figures/traveling_wave}
    \includegraphics[width=0.48\textwidth]{figures/waves_D}
    \includegraphics[width=0.48\textwidth]{figures/waves_v}
    \caption{\label{fig:traveling_wave} {\bf Phylogeographic estimates in rapidly changing environments.} Panel A shows an illustration of the simulated set-up: a carrying capacity with a Gaussian profile with $\sigma=0.5$ that moves along the $x$-axis with velocity $v$ on a patch with $L_x=3$, $L_y=1$, and periodic boundary conditions.
    Panels B\&C show estimates of diffusion coefficients and velocities, relative to their true values, as a function of the ratio of the FKPP velocity to the external velocity $v$. }
\end{figure}

\begin{itemize}
    \item periodic seasaws and other patters
    \item can we trust ancestral location inferences? with true D, with estimated D? significant biases?
    \item can we say something general? coalescence time versus diffusion time across the habitat, habitat changes vs fkpp speed
    \item typical FKPP speed: $D=km^2/day$, $\alpha = 0.2/day$. $v_{FKPP} = 2\sqrt{D\alpha} = km/day$.
\end{itemize}

\section*{Discussion}
